Towards a Wide-Coverage Grammar for Swedish\\ Using GF

### 2012-01-13 doc/report/report.tex
planning

\abstract{
The goal of this thesis is to enable open-domain parsing for Swedish
by using the GF \\grammar formalism,
%by %The idea has been to% 
combining existing language technology resources with
new techniques.
The work includes the development of a covering grammar,
%in the GF formalism
to identify and formalize \\
grammatical constructions %specific to and
frequent in Swedish. % have been investigated and formalised in the functional
%programming language.
We compare and evaluate the possibilities of using GF
for modelling large-scale grammars and for 
%and the aim to reaching our goal;
obtaining a system for analysing written text.
To reach this goal, problems of computational as well as linguistic nature had to be solved.
%~~achieve a covering Swedish~~ grammar implemented
%TODO first sentence should be more, no understatements!
%in GF -- to model natural language by functional programming. 
\\We present the methods and technologies used and further inspect the possibilities
of reaching our long-term goal; to accomplish a parser for unrestricted Swedish, by
combining our rule-based grammar with statistical methods.
\\Our contribution is a wide-coverage GF lexicon, a translation of a Swedish treebank
to GF notation and an extended Swedish grammar implementation. The grammar is based
on the multilingual abstract syntax given in GF resource library, and now
also covers constructions specific to Swedish.
We further give an example of the advantage of using dependent types when
describing grammar and syntax, in this case for dealing with reflexive pronouns.
%the linguistic phenomena

@@

Parse trees on the other hand show only the types assigned to the words and phrases.
Information about tense, polarity etc. which are explicitly given in the abstract tree are
not reproduced in the parse tree. Hence, parse trees do not give complete representations
but model the parse resluts in a transparent manner. For our example, the definiteness and
number of the noun `\emph{katten}' is shown as \verb-DetQuant DefArt NumSg- in
the abstract tree while the parse tree only show that a the noun phrase consists of one noun. 
% is used but not in the parse tree.
In the corresponding English parse tree, figure \ref{pic:gfEtree},
the noun is explicitly quantified by the article \emph{`the'},
and the determiner, the first argument to the function \verb-DetCN-, is
therefore shown in the parse tree.\\

@@

The semantic aspect of Saldo requires that words with multiple meanings are
separated into different entries.
%TODO change word? uppskatta, känna, få
The word `uppskatta' for example, has two entries, with sligthly
different semantics: \emph{enjoy}
\enumsentence{Jag uppskattar teater. \\
``I enjoy theater."}
or \emph{estimate}. 
\enumsentence{Värdet uppskattades till 5 miljoner.\\
``The value was estimated to 5 million."}
%\enumsentence{Barnen växte varje dag\\
%``The children grew for each day"}
%or \emph{vegetate}. 
%\enumsentence{Det växte en gammal ek där.\\
%``There stood an old oak."}
`uppskatta' is however declined the same way for both interpretations of the word.
Hence \textit{Saldo's morphological lexicon} only keeps one entry of `uppskatta'.
For the homonyms that do not share the whole inflection tables, different entries are 
kept also in the morphological lexicon.\\
For our purpose, only the morphological section, provided under LPGL in XML
format, was needed.
The data can be processed and translated to GF format as described in 
section \ref{sec:prog.saldo}.

@@

%\enumsentence{\shortexnt{8}
%{Har& han& inte& ätit& de & gula& äpplena& idag?}
%{Has& he& not& eaten& the& yellow& apples& today?}} 
%\label{gfSwe:parseable}
%Covers sentences like
% better sentence where the verb phrase is split up?
\begin{figure}[h]
\includegraphics[width=70mm,height=70mm]{apples1.pdf}
\caption{Parse tree for ``Har han inte ätit de gula äpplena idag?"}
\label{gfSwe:parsetree}
\end{figure}
Even though the verb phrase \emph{``har inte ätit de gula äpplena idag"} is discontinuous, the whole
phrase is still treated as one constituent in GF. The parts
are connected in the tree, and the subject \emph{`han'} is put between the
finite verb and the rest of the phrase.
At the code level, this is done using the record type for \verb-VP-, wich
consists of fields that can be put %the \verb-VP- fields 
in different order.\\
\verb|table { Inv => verb.fin ++ subj ++ verb.neg ++ verb.inf ++ verb.compl ; ... | \\
%At the code level, this is implemented %using records.
%The linearization rule for questions -- clauses with inverted order -- picks 
%the field for the finite verb and appends it to the subject followed by
%the negation, infinite part and the complement of the verb phrase: \\

@@

%The mapping works for Talbanken05 in Tiger XML-format.
\section{The Talbanken annotation}
Talbanken05 uses three kinds of tags: categories, edge labels and POS-tags. 
While the POS-tags are reserved for words, the categories give grammatictal information,
such as \verb|S|, \verb|NP| and \verb|VP|.
The edge labels give the
part of sentence: \verb|SS| for subject, 
\verb|OO| for object etc. Nearly 400 different tags are used, of which more than
300 different are POS-tags. The high number is due to the level of detail,  there are for
example almost 50 tags for nouns, excluding proper names. The tags
show definiteness, case and whether the word is a compound. Some words also have
their own tags.\\

@@

\chapter{Discussion}
\label{sec:results}
\section{Evaluation}
%TODO
%>>evolution diagram?
%also removed unnecessary ambigutiies.
%>> how to handle ambigs: know them and controll them. using statistics or ontolgies.
%>> that is: disambiguitation by context or by meaning.
The project has resulted in
\begin{itemize}
\item a large-scale GF lexicon and a program to redo the importation when needed
\item an extended grammar covering an importand part of Swedish
\item a comparison between GF and another annotation
\item a deeper testing of the Swedish resource grammar and an estimation
of how well GF can be used to describe larger parts of a language
\item a study of how dependent types can be used in the resource grammars
\end{itemize}

@@

The grammar has been extended and enhanced, and its current status is
a specialized extension of the resource grammar.
Besides parsing, the grammar may well be used for language generation,
which works fast even when using an extensive lexicon.
%TODO make nice!
Although it is not been formally verified, we believe that the majority of the
sentences generated are grammatically correct in a syntactical point of view.
Without any semantic knowledge, nonsense phrases cannot be avoided in
random generation. However, given that the abstract tree has a meaningful
interpretation,
%TODO move to Misc?
the linearization should be correct. There are some cases when the correctness
of the output has been put aside in order to increase the expressivity, such
as \verb-ComplBareVV-, which use a verb-complement verb without the infinitive mark, 
as in sentence (\ref{sent:complVV1}b).
\eenumsentence{\item\shortex{5}
{Jag& börjar&\textbf{att}& bli& hungrig}
{\emph{I}& \emph{begin}& \textbf{\emph{to}}&\emph{become}& \emph{hungry}}
{``I'm getting hungry"}
\item
\shortex{4}
{Jag& börjar& bli& hungrig}
{\emph{I}& \emph{begin}& \emph{become}& \emph{hungry}}
{``I'm getting hungry"}}\label{sent:complVV1}
Since there is no information in the grammar or lexicon about which verbs that
allows the infinitive mark to be left out, it will also allow the more
questionable sentence (\ref{sent:complVV2}b).
\eenumsentence{
\item \shortex{5}
{Jag & låter &bli & \textbf{att}& titta} 
{\emph{I} & \emph{let} & \emph{be} &\textbf{\emph{to}} &\emph{watch}}
{``I refrain from watching"}
\item \shortexnt{4}
{*Jag & låter & bli & titta} 
{~~\emph{I} & \emph{leave} &\emph{be} & \emph{watch}}}\label{sent:complVV2}
As these functions serve to provide stylistic flexiblity when parsing, they can be left out
when generating language, and the grammar then generates text of good
grammatical quality. \\
%ComplBareVV, SupCl
%For random generation, the sentences generated may be next to impossible
%to interpret without looking at the abstract tree, since they may contain
%too many attributes, relative clauses etc.

>>>>

\abstract{
This thesis describes the work to develop a wide-coverage grammar for
parsing and generating Swedish text.
We do this by using the dependently typed grammar formalism GF,
a functional programming language specialized at describing grammars.
The idea is to 
%The goal of this thesis is to enable open-domain parsing for Swedish
%by using the GF \\grammar formalism,
%by %The idea has been to% 
combine existing language technology resources with
new techniques and our motivation is the goal of achieving a parser
for unrestricted Swedish.
To reach this goal, problems of computational as well as linguistic nature had
to be solved.  The work includes the development of the grammar,
%in the GF formalism
to identify and formalize 
grammatical constructions %specific to and
frequent in Swedish, as well as methods for importing a large-scale lexicon and to evaluate
the parser. % have been investigated and formalised in the functional
%programming language.
%~~achieve a covering Swedish~~ grammar implemented
%and the aim to reaching our goal;
%TODO first sentence should be more, no understatements!
%in GF -- to model natural language by functional programming. 
\\We present the methods and technologies used and further inspect the possibilities
of reaching our long-term goal; to accomplish a parser for unrestricted Swedish, by
combining our rule-based grammar with statistical means.
We compare and discuss the advantages and problems of using GF
for modeling large-scale grammars.
%and for obtaining a system for analysing written text.
\\Our contribution is a wide-coverage GF lexicon, a translation of a Swedish treebank
to GF notation and an extended Swedish grammar implementation. The grammar is based
on the multilingual abstract syntax given in GF resource library, and now
also covers constructions specific to Swedish.
We further give an example of the advantage of using dependent types when
describing grammar and syntax, in this case for dealing with reflexive pronouns.
%the linguistic phenomena

@@

Parse trees on the other hand show only the types assigned to the words and phrases.
Information about tense, polarity etc. which are explicitly given in the abstract tree are
not reproduced in the parse tree. Hence, parse trees do not give complete representations
but model the parse results in a transparent manner. For our example, the definiteness and
number of the noun `\emph{katten}' is shown as \verb-DetQuant DefArt NumSg- in
the abstract tree while the parse tree only show that a the noun phrase consists of one noun. 
% is used but not in the parse tree.
In the corresponding English parse tree, figure \ref{pic:gfEtree},
the noun is explicitly quantified by the article \emph{`the'},
and the determiner, the first argument to the function \verb-DetCN-, is
therefore shown in the parse tree.\\

@@

The semantic aspect of Saldo requires that words with multiple meanings are
separated into different entries.
%TODO change word? uppskatta, känna, få
The word `uppskatta' for example, has two entries, with slightly
different semantics: \emph{enjoy}
\enumsentence{Jag uppskattar teater. \\
``I enjoy theater."}
or \emph{estimate}. 
\enumsentence{Värdet uppskattades till 5 miljoner.\\
``The value was estimated to 5 million."}
%\enumsentence{Barnen växte varje dag\\
%``The children grew for each day"}
%or \emph{vegetate}. 
%\enumsentence{Det växte en gammal ek där.\\
%``There stood an old oak."}
`uppskatta' is however declined the same way for both interpretations of the word.
Hence \textit{Saldo's morphological lexicon} only keeps one entry of `uppskatta'.
For the homonyms that do not share the whole inflection tables, different entries are 
kept also in the morphological lexicon.\\
For our purpose, only the morphological section, provided under LPGL in XML
format, was needed.
The data can be processed and translated to GF format as described in 
section \ref{sec:prog.saldo}.

@@

%\enumsentence{\shortexnt{8}
%{Har& han& inte& ätit& de & gula& äpplena& idag?}
%{Has& he& not& eaten& the& yellow& apples& today?}} 
%\label{gfSwe:parseable}
%Covers sentences like
% better sentence where the verb phrase is split up?
\begin{figure}[h]
\includegraphics[width=70mm,height=70mm]{apples1.pdf}
\caption{Parse tree for ``Har han inte ätit de gula äpplena idag?"}
\label{gfSwe:parsetree}
\end{figure}
Even though the verb phrase \emph{``har inte ätit de gula äpplena idag"} is discontinuous, the whole
phrase is still treated as one constituent in GF. The parts
are connected in the tree, and the subject \emph{`han'} is put between the
finite verb and the rest of the phrase.
At the code level, this is done using the record type for \verb-VP-, which
consists of fields that can be put %the \verb-VP- fields 
in different order.\\
\verb|table { Inv => verb.fin ++ subj ++ verb.neg ++ verb.inf ++ verb.compl ; ... | \\
%At the code level, this is implemented %using records.
%The linearization rule for questions -- clauses with inverted order -- picks 
%the field for the finite verb and appends it to the subject followed by
%the negation, infinite part and the complement of the verb phrase: \\

@@

%The mapping works for Talbanken05 in Tiger XML-format.
\section{The Talbanken annotation}
Talbanken05 uses three kinds of tags: categories, edge labels and POS-tags. 
While the POS-tags are reserved for words, the categories give grammatical information,
such as \verb|S|, \verb|NP| and \verb|VP|.
The edge labels give the
part of sentence: \verb|SS| for subject, 
\verb|OO| for object etc. Nearly 400 different tags are used, of which more than
300 different are POS-tags. The high number is due to the level of detail,  there are for
example almost 50 tags for nouns, excluding proper names. The tags
show definiteness, case and whether the word is a compound. Some words also have
their own tags.\\

@@

\chapter{Discussion}
\label{sec:results}
\section{Evaluation}
%TODO
%>>evolution diagram?
%also removed unnecessary ambigutiies.
%>> how to handle ambigs: know them and controll them. using statistics or ontolgies.
%>> that is: disambiguitation by context or by meaning.
The project has resulted in
\begin{itemize}
\item a large-scale GF lexicon and a program to redo the importation when needed
\item an extended grammar covering an important part of Swedish
\item a comparison between GF and another annotation
\item a deeper testing of the Swedish resource grammar and an estimation
of how well GF can be used to describe larger parts of a language
\item a study of how dependent types can be used in the resource grammars
\end{itemize}

@@

The grammar has been extended and enhanced, and its current status is
a specialized extension of the resource grammar.
Besides parsing, the grammar may well be used for language generation,
which works fast even when using an extensive lexicon.
%TODO make nice!
Although it is not been formally verified, we believe that the majority of the
sentences generated are grammatically correct in a syntactical point of view.
Without any semantic knowledge, nonsense phrases cannot be avoided in
random generation. However, given that the abstract tree has a meaningful
interpretation,
%TODO move to Misc?
the linearization should be correct. There are some cases when the correctness
of the output has been put aside in order to increase the expressivity, such
as \verb-ComplBareVV-, which use a verb-complement verb without the infinitive mark, 
as in sentence (\ref{sent:complVV1}b).
\eenumsentence{\item\shortex{5}
{Jag& börjar&\textbf{att}& bli& hungrig}
{\emph{I}& \emph{begin}& \textbf{\emph{to}}&\emph{become}& \emph{hungry}}
{``I'm getting hungry"}
\item
\shortex{4}
{Jag& börjar& bli& hungrig}
{\emph{I}& \emph{begin}& \emph{become}& \emph{hungry}}
{``I'm getting hungry"}}\label{sent:complVV1}
Since there is no information in the grammar or lexicon about which verbs that
allows the infinitive mark to be left out, it will also allow the more
questionable sentence (\ref{sent:complVV2}b).
\eenumsentence{
\item \shortex{5}
{Jag & låter &bli & \textbf{att}& titta} 
{\emph{I} & \emph{let} & \emph{be} &\textbf{\emph{to}} &\emph{watch}}
{``I refrain from watching"}
\item \shortexnt{4}
{*Jag & låter & bli & titta} 
{~~\emph{I} & \emph{leave} &\emph{be} & \emph{watch}}}\label{sent:complVV2}
As these functions serve to provide stylistic flexibility when parsing, they can be left out
when generating language, and the grammar then generates text of good
grammatical quality. \\
%ComplBareVV, SupCl
%For random generation, the sentences generated may be next to impossible
%to interpret without looking at the abstract tree, since they may contain
%too many attributes, relative clauses etc.
